GEE models

### 2016-06-28 doc/manuscript.Rmd
Updated analysis and manuscript as per Adria's suggestions

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
output: fost::diabetes_care_manuscript
---

@@

Word count: 3577 / 4000

@@

# Abstract
248 / 250 words

@@

Research Design and Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and who had completed an OGTT at 3 time
points over 6 years. Outcome variables were calculated using the OGTT values.
IS was assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function
was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, age, ALT, alcohol intake,
physical activity, and family history of diabetes. NEFA were analyzed as both
concentrations and proportions (mol%) of the total fraction.

@@

```{r knit_setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'), dpi = 180)
```

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models). Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')

@@

sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with blue indicating positive and orange indicating negative correlations.')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. Models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')
stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models). Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, indicated by asterisk.')

@@

sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')

@@

sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis.')
sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA). Discriminatory ability is evident from the amount of separation between the two groups. In this case, PLS-DA was 82% accurate at correctly classifying groups.')
sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).  The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and in addition can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a].
Observational and clinical studies have reported concordant findings, showing
in particular that elevated total plasma NEFA associates with an increased risk
for incident type 2 diabetes [@Daniele2014a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple molecules with diverse physiological functions, and few 
studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
been epidemiological studies that only looked at total NEFA and not individual
fatty acids [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical to
understanding associations at a more granular level given the protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual fatty acids. Therefore, our objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given their
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see [@Johnston2016a] for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given its capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last four models (M5-M8) were comparable in performance, therefore we
report results from the simplest model (M8), adjusting for time, sex, ethnicity,
baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While we decided *a priori*
to test for and report on the time by predictor interaction results, we
also assessed the overall fit of models that included a time interaction, which
ultimately did not improve the overall fit of our models in terms of QIC
(M6.1 and M8.1 as seen
in `r cite_st('qic')`) and we therefore did not include them in the main effect
analysis. Given that TAG is a risk factor
for diabetes and that NEFA contribute to TAG production, TAG may act as a mediator
between NEFA and the outcomes. To determine the role of TAG in the
association between NEFA and the outcomes, TAG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. As such, after running the GEE models for the
longitudinal analysis, partial least squares discriminant analysis (PLS-DA) was
used to identify the patterns of NEFA composition against beta-cell function
(using ISSI-2, which had the majority of the significant associations in the GEE models) and for
those who converted to or maintained dysglycemia status (either IFG, IGT, or DM)
over the 6 years. Since PLS-DA cannot handle longitudinal data, latent class
mixed models (LCMM) were used to extract latent trajectories of beta-cell
function (ISSI-2) over the 6 years.

@@

outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(gee_df), value = TRUE)
ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
```

```{r geeData, cache=TRUE}
gee_results <- dplyr::bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()

gee_results_unadj <- dplyr::bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()
```

@@

For this study, there were 
`r sex.npct$npct` females and `r ethn.npct$npct` had European-ancestry.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r fhd.npct$npct`, had a family history of diabetes.
The primary outcome variables showed a significant median decline between 
`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
years in this analysis of the PROMISE cohort.
Consistent with this decline, there were `r dm.convert`
participants who developed diabetes while `r predm.convert` participants either
converted to or maintained IFG or IGT status over the 6 years.

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
Full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
had negative associations with
IGI/IR and ISSI-2. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD increase
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. However, in contrast to these results using NEFA modeled as concentrations, none of the NEFA variables modeled as mol% were associated with the
outcomes. Adjusting for TAG attenuated all associations with IGI/IR and ISSI-2
(data not shown).
There were no significant interaction effects
between time, sex, or ethnicity and the individual NEFA on any of the outcome
measures (all p>0.15). Raw values from the GEE models are shown in 
`r cite_st('gee')` for fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.

## Latent trajectory and NEFA pattern recognition

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
#csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
csl: /home/luke/Documents/bibliography/styles/vancouver-brackets-no-et-al.csl
output: fost::diabetes_care_manuscript
---

@@

Word count: 3543 / 4000

@@

# Abstract
243 / 250 words

@@

Research Design and Methods: In the PROMISE cohort, 477 participants had serum
NEFA measured at the baseline visit and completed an OGTT at 3 time
points over 6 years. Outcome variables were calculated using the OGTT values.
IS was assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function
was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
used, adjusting for time, waist, sex, ethnicity, baseline age, ALT, and family
history of diabetes. NEFA were analyzed as both concentrations and proportions
(mol%) of the total fraction.

@@

```{r knit_setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '../')
```

@@

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')

@@

sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort. GEE models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, presented as the dot size.')
stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. GEE models were adjusted for time, sex, ethnicity, baseline age, WC,
ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equation models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  GEE models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')

@@

sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%. See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot.')

@@

sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.')
sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA). Discriminatory ability is evident from the amount of separation between the two groups. In this case, PLS-DA was 82% accurate at correctly classifying groups. See the Supplemental Methods for a description of PLS-DA and for interpreting this plot.')
sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).  See the Supplemental Methods for a description of PLS-DA and interpreting this plot. The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and in addition can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a].
Observational and clinical studies have reported concordant findings, showing
in particular that elevated total plasma NEFA associates with an increased risk
for incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple molecules with diverse physiological functions, and few 
studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Kato2008].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in this research. The majority of previous 
studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], 
have been short term human trials 
[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have
been epidemiological studies that only looked at total NEFA and not individual
fatty acids [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical given the protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual fatty acids. Therefore, our objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower PUFA such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given their
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see ref @Johnston2016a for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC. While the
final GEE model selected as best fitting differed between insulin sensitivity
and beta-cell function measures, the differences in QIC values were less than
10 between the models, suggesting similar fit. As such, we selected the model
that had the fewest covariates and that had similar fit between the outcome
measures (M8). The final GEE model was adjusted for time, sex, ethnicity,
baseline age, WC, ALT, and family history of diabetes. After scaling,
log-transforming, and exponentiating, the GEE estimates are
interpreted as an expected percent difference in the outcome variable for every
SD increase in the predictor variable given the covariates are held constant.
Since TAG is a risk factor for diabetes and since NEFA contribute to TAG
production, TAG may act as a mediator between NEFA and the outcomes. To
determine the role of TAG in the association between NEFA and the outcomes, TAG
was included in the GEE model as a sensitivity analysis. Lastly for the GEE
models, we tested for an interaction with sex, ethnicity, or time by the
predictor term for each outcome variable.

While GEE accounts for the longitudinal design of the data, this approach is
limited in that it cannot analyze the inherent multivariate nature of the
composition of the NEFA fraction. Therefore, to confirm the results of the GEE
analyses in a multivariate environment (all NEFA variables in a single model),
partial least squares discriminant analysis (PLS-DA) was used to identify the
patterns of NEFA composition against beta-cell function (using ISSI-2, which had
the majority of the associations in the GEE models) and for those
who converted to or maintained dysglycemia status (either IFG, IGT, or DM) over
the 6 years. Latent class mixed models (LCMM) were used to extract the underlying
trajectories of beta-cell function (ISSI-2) over the 6 years. For more detailed
explanation of GEE, PLS-DA, and LCMM, please see the Supplemental Methods.

@@

outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(gee_df), value = TRUE)
ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
covariates <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
```

```{r geeData, cache=TRUE}
gee_results <- dplyr::bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates, 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()

gee_results_unadj <- dplyr::bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()
```

@@

For this study, there were 
`r sex.npct$npct` females and `r ethn.npct$npct` had European-ancestry.
The average age of the participants was `r aide::average(ds0yr$Age)` years 
(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r fhd.npct$npct`, had a family history of diabetes.
Insulin sensitivity and beta-cell function measures showed a significant median
decline between
`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
years in this analysis of the PROMISE cohort.
Consistent with this decline, there were `r dm.convert`
participants who developed diabetes while `r predm.convert` participants either
converted to or maintained IFG or IGT status over the 6 years.
<!-- maintain vs convert -->

@@

A number of associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
While full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
continued to show negative associations with IGI/IR and ISSI-2. Sensitivity
analyses revealed that adjustment for WC was responsible for attenuating the
unadjusted results. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD increase
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. However, in contrast to these results using NEFA modeled as
concentrations, none of the NEFA variables modeled as mol% were associated with
the outcomes. Adjusting for TAG attenuated all associations with IGI/IR and
ISSI-2 (data not shown). Sensitivity analyses to examine the association with
the sum of fatty acid classes (saturated vs unsaturated) revealed similar
findings as seen with the adjusted GEE model for total NEFA (data not shown).
There were no significant interaction effects between time, sex, or ethnicity
and the individual NEFA on any of the outcome measures, except for a single
marginally significant (p=0.04) interaction between sex, 20:3n-6 (mol%), and
ISI, which showed that males had a 12% higher ISI for every SD increase in
20:3n-6. Raw values from the GEE models are shown in `r cite_st('gee')` for
fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.

## Latent trajectories and differences in NEFA composition
